Automated Residual Plot Assessment with the R Package autovi and the Shiny App autovi.web

### 2025-05-12 autovi_paper.qmd
Refine lengthy sentences

---
title: "Automated Residual Plot Assessment with the R Package autovi and the Shiny App autovi.web"
format:
  quarto-anzjs-pdf:
    keep-tex: true  
    shorttitle: "Automated Residual Plot Assessment"
    double-space: true
    line-numbers: true
    # You can specify extra preamble in the file below.
    include-in-header:
      - header.tex
date: today
date-format: "YYYY" # used to define volumeyear
author:
  - name: Weihao Li
    email: weihao.li@monash.edu
    corresponding: true
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0003-4959-106X
  - name: Dianne Cook
    email: dicook@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0002-3813-7155
  - name: Emi Tanaka
    email: emi.tanaka@anu.edu.au
    affiliations: 
      - name: The Australian National University
        department: Biological Data Science Institute
        address: 46 Sullivan's Creek Road, ACT 2600, Australia
    orcid: 0000-0002-1455-259X
  - name: Susan VanderPlas
    email: emi.tanaka@anu.edu.au
    affiliations:
      - name: University of Nebraska
        department: Department of Statistics
        address: Hardin Hall, 3310 Holdrege St Suite 340, Lincoln, NE 68583, United States
    orcid: 0000-0002-3803-0972
  - name: Klaus Ackermann
    email: Klaus.Ackermann@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0001-7693-8538
# Note that the Journal requires that a paper must begin with a
# "Summary" not an "Abstract".  This is automatically taken care
# of by the anzsauth document style.  So even though the following
# says "abstract" the heading "Summary" will appear in
# the processed version.
abstract: |
  Visual assessment of residual plots is a common approach for diagnosing linear models, but it relies on manual evaluation, which does not scale well and can lead to inconsistent decisions across analysts. The lineup protocol, which embeds the observed plot among null plots, can reduce subjectivity but requires even more human effort. In today’s data-driven world, such tasks are well-suited for automation. We present a new R package that uses a computer vision model to automate the evaluation of residual plots. An accompanying Shiny app is provided for ease of use. Given a sample of residuals, the model predicts a visual signal strength (VSS) and offers supporting information to help analysts assess the adequacy of their model fit. 
# Note that "keywords" should not include words and phrases
# that form part of the title of the paper.
# Also keywords (with the exception of proper names and
# certain abbreviations that conventionally appear all in
# capital letters) should *not* be capitalised.
keywords: [initial data analysis, statistical graphics, data visualization, visual inference, computer vision, machine learning, hypothesis testing, regression analysis, model diagnostics]
bibliography: bibliography.bib  
#nocite: |
---

@@

VSS is an estimate of the distance between the residual distribution of a fitted classical normal linear regression model and a reference distribution; more details can be found in @li2024automated. The distance measure is based on the Kullback-Leibler (KL) divergence:

@@

The third example is generated using the `surreal` package [@surreal] where structured residuals are hidden in data, to be revealed if the correct model is specified. Here a quote based on Tukey is used as the residual structure "visual summaries focus on unexpected values". The automated plot reading predicts the VSS to be 5.87, with a $p$-value less than 0.05. This structure is blindingly obvious visually, but a RESET test for nonlinear structure would not report a problem. (It would be detected by a Breusch-Pagan for heteroscedasticity and also Shapiro-Wilk test [@shapiro1965analysis] for non-normality.)

@@

Due to the resource allocation policy of shinyapps.io, the server enters a sleep mode during periods of inactivity, resulting in the clearing of the local Python virtual environment. Consequently, when the application "wakes up" for a new user session, these libraries need to be reinstalled. While this ensures a clean environment for each session, it may lead to slightly longer loading times for the first user after a period of inactivity.

>>>>

---
title: "Automated Residual Plot Assessment with the R Package autovi and the Shiny App autovi.web"
format:
  quarto-anzjs-pdf:
    keep-tex: true  
    shorttitle: "Automated Residual Plot Assessment"
    double-space: true
    line-numbers: true
    # You can specify extra preamble in the file below.
    include-in-header:
      - header.tex
date: today
date-format: "YYYY" # used to define volumeyear
author:
  - name: Weihao Li
    email: weihao.li@monash.edu
    corresponding: true
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0003-4959-106X
  - name: Dianne Cook
    email: dicook@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0002-3813-7155
  - name: Emi Tanaka
    email: emi.tanaka@anu.edu.au
    affiliations: 
      - name: The Australian National University
        department: Biological Data Science Institute
        address: 46 Sullivan's Creek Road, ACT 2600, Australia
    orcid: 0000-0002-1455-259X
  - name: Susan VanderPlas
    email: emi.tanaka@anu.edu.au
    affiliations:
      - name: University of Nebraska
        department: Department of Statistics
        address: Hardin Hall, 3310 Holdrege St Suite 340, Lincoln, NE 68583, United States
    orcid: 0000-0002-3803-0972
  - name: Klaus Ackermann
    email: Klaus.Ackermann@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0001-7693-8538
# Note that the Journal requires that a paper must begin with a
# "Summary" not an "Abstract".  This is automatically taken care
# of by the anzsauth document style.  So even though the following
# says "abstract" the heading "Summary" will appear in
# the processed version.
abstract: |
  Visual assessment of residual plots is a common approach for diagnosing linear models, but it relies on manual evaluation, which does not scale well and can lead to inconsistent decisions across analysts. The lineup protocol, which embeds the observed plot among null plots, can reduce subjectivity but requires even more human effort. In today’s data-driven world, such tasks are well-suited for automation. We present a new R package that uses a computer vision model to automate the evaluation of residual plots. An accompanying Shiny app is provided for ease of use. Given a sample of residuals, the model predicts a visual signal strength (VSS) and offers supporting information to help analysts assess model fit. 
# Note that "keywords" should not include words and phrases
# that form part of the title of the paper.
# Also keywords (with the exception of proper names and
# certain abbreviations that conventionally appear all in
# capital letters) should *not* be capitalised.
keywords: [initial data analysis, statistical graphics, data visualization, visual inference, computer vision, machine learning, hypothesis testing, regression analysis, model diagnostics]
bibliography: bibliography.bib  
#nocite: |
---

@@

VSS estimates the distance between the residual distribution of a fitted classical normal linear regression model and a reference distribution [see @li2024automated, for details]. The distance measure is based on the Kullback-Leibler (KL) divergence:

@@

The third example is generated using the `surreal` package [@surreal], where structure residuals are embedded in the data. In this case, a quote inspired by Tukey, "visual summaries focus on unexpected values", is used to define the residual structure. The automated plot reading predicts the VSS to be 5.87, with a $p$-value less than 0.05. Visually, the structure is strikingly clear, but a RESET test for nonlinear structure would not report a problem. (It would be detected by a Breusch-Pagan for heteroscedasticity and also Shapiro-Wilk test [@shapiro1965analysis] for non-normality.)

@@

Due to shinyapps.io's resource policy, inactive servers enter sleep mode, clearing the local Python environment. When reactivated for a new session, libraries must be reinstalled. While this ensures a clean environment for each session, it may lead to slightly longer loading times for the first user after a period of inactivity.

### 2025-05-12 autovi_paper.tex
Refine lengthy sentences

\begin{abstract}
Visual assessment of residual plots is a common approach for diagnosing
linear models, but it relies on manual evaluation, which does not scale
well and can lead to inconsistent decisions across analysts. The lineup
protocol, which embeds the observed plot among null plots, can reduce
subjectivity but requires even more human effort. In today's data-driven
world, such tasks are well-suited for automation. We present a new R
package that uses a computer vision model to automate the evaluation of
residual plots. An accompanying Shiny app is provided for ease of use.
Given a sample of residuals, the model predicts a visual signal strength
(VSS) and offers supporting information to help analysts assess the
adequacy of their model fit.
\end{abstract}

@@

VSS is an estimate of the distance between the residual distribution of
a fitted classical normal linear regression model and a reference
distribution; more details can be found in \citet{li2024automated}. The
distance measure is based on the Kullback-Leibler (KL) divergence:

@@

The third example is generated using the \texttt{surreal} package
\citep{surreal} where structured residuals are hidden in data, to be
revealed if the correct model is specified. Here a quote based on Tukey
is used as the residual structure ``visual summaries focus on unexpected
values''. The automated plot reading predicts the VSS to be 5.87, with a
\(p\)-value less than 0.05. This structure is blindingly obvious
visually, but a RESET test for nonlinear structure would not report a
problem. (It would be detected by a Breusch-Pagan for heteroscedasticity
and also Shapiro-Wilk test \citep{shapiro1965analysis} for
non-normality.)

@@

Due to the resource allocation policy of shinyapps.io, the server enters
a sleep mode during periods of inactivity, resulting in the clearing of
the local Python virtual environment. Consequently, when the application
``wakes up'' for a new user session, these libraries need to be
reinstalled. While this ensures a clean environment for each session, it
may lead to slightly longer loading times for the first user after a
period of inactivity.

>>>>

\begin{abstract}
Visual assessment of residual plots is a common approach for diagnosing
linear models, but it relies on manual evaluation, which does not scale
well and can lead to inconsistent decisions across analysts. The lineup
protocol, which embeds the observed plot among null plots, can reduce
subjectivity but requires even more human effort. In today's data-driven
world, such tasks are well-suited for automation. We present a new R
package that uses a computer vision model to automate the evaluation of
residual plots. An accompanying Shiny app is provided for ease of use.
Given a sample of residuals, the model predicts a visual signal strength
(VSS) and offers supporting information to help analysts assess model
fit.
\end{abstract}

@@

VSS estimates the distance between the residual distribution of a fitted
classical normal linear regression model and a reference distribution
\citep[see][for details]{li2024automated}. The distance measure is based
on the Kullback-Leibler (KL) divergence:

@@

The third example is generated using the \texttt{surreal} package
\citep{surreal}, where structure residuals are embedded in the data. In
this case, a quote inspired by Tukey, ``visual summaries focus on
unexpected values'', is used to define the residual structure. The
automated plot reading predicts the VSS to be 5.87, with a \(p\)-value
less than 0.05. Visually, the structure is strikingly clear, but a RESET
test for nonlinear structure would not report a problem. (It would be
detected by a Breusch-Pagan for heteroscedasticity and also Shapiro-Wilk
test \citep{shapiro1965analysis} for non-normality.)

@@

Due to shinyapps.io's resource policy, inactive servers enter sleep
mode, clearing the local Python environment. When reactivated for a new
session, libraries must be reinstalled. While this ensures a clean
environment for each session, it may lead to slightly longer loading
times for the first user after a period of inactivity.
